---
title: "Assignment 5 - Decision Trees"
author: "Charles Lang"
date: "November 9, 2016"
output: html_document
---

For this assignment we will be using data from the Assistments Intelligent Tutoring system. This system gives students hints based on how they perform on math problems. 

#Install & call libraries
```{r}
install.packages("party", "rpart")

library(rpart)
library(party)
```

## Part I
```{r}
D1 <- read.csv("intelligent_tutor.csv")
```

##Classification Tree
First we will build a classification tree to predict which students ask a teacher for help, which start a new session, or which give up, based on whether or not the student completed a session (D1$complete) and whether or not they asked for hints (D1$hint.y). 
```{r}

c.tree <- rpart(action ~ hint.y + complete, method="class", data=D1) #Notice the standard R notion for a formula X ~ Y

#Look at the error of this tree
printcp(c.tree)

#Plot the tree
post(c.tree, file = "tree.ps", title = "Session Completion Action: 1 - Ask teacher, 2 - Start new session, 3 - Give up")

```
## Part II

#Regression Tree

We want to see if we can build a decision tree to help teachers decide which students to follow up with, based on students' performance in Assistments. We will create three groups ("teacher should intervene", "teacher should monitor student progress" and "no action") based on students' previous use of the system and how many hints they use. To do this we will be building a decision tree using the "party" package. The party package builds decision trees based on a set of statistical stopping rules.

#Take a look at our outcome variable "score"
```{r}
hist(D1$score)
```

#Create a categorical outcome variable based on student score to advise the teacher using an "ifelse" statement
```{r}
D1$advice <- ifelse(D1$score <=0.4, "intervene", ifelse(D1$score > 0.4 & D1$score <=0.8, "monitor", "no action"))
```

#Build a decision tree that predicts "advice" based on how many problems students have answered before, the percentage of those problems they got correct and how many hints they required
```{r}
score_ctree <- ctree(factor(advice) ~ prior_prob_count + prior_percent_correct + hints, D1)
#score_ctree <- ctree(factor(advice) ~ ., D1)
```

#Plot tree
```{r}
plot(score_ctree)
```

##Please interpret the tree, which two behaviors do you think the teacher should most closely pay attemtion to?
Students who asked for more than 12 hints, or students who asked for asked for some hints (but less than 12) and got less than 63% correct in their prior score.

#Test Tree
Upload the data "intelligent_tutor_new.csv". This is a data set of a different sample of students doing the same problems in the same system. We can use the tree we built for the previous data set to try to predict the "advice" we should give the teacher about these new students. 

```{r}
#Upload new data

D2 <- read.csv("intelligent_tutor_new.csv")

#Generate predicted advice for new students based on tree generated from old students

D2$prediction <- predict(score_ctree, D2)
D2$hint.y <- with(D2, ifelse(hints > 0, 0, 1))
table(D2$prediction)
``` 

Based on the previous tree, the teachers should monitor 84 situations, intervene in none and take no action in 116


## Part III
Compare the predicted advice with the actual advice that these studnts recieved. What is the difference between the observed and predicted results?

The score is the actual scores students received.  All students receive all answers correctly, so in this particular case, the model suggest that teachers monitor 84 cases where there is no need to take an action (because they actually received 100%).

